---
title: "Assignment 10: Data Scraping"
author: "Alejandra Rios"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
getwd()

library(tidyverse)
library(lubridate)
library(viridis)
library(here)
library(rvest)
library(dataRetrieval)
library(dplyr)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
webpage <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")
webpage


```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
system_name

pwsid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pwsid

ownership <- webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
ownership

max_mgd <- webpage %>%
  html_nodes("th~ td+ td") %>% 
  html_text()
max_mgd
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
year <- "2022"

#scrape month
month <- webpage %>% html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% html_text()

#make withdrawls dataframe
withdrawls <- data.frame("Month" = c(month),"Year" = c(year),"Date" = my(paste(month,"-",year)),
                                                                         "MGD" = as.numeric(max_mgd),
                                                                         "PWSID" = rep(pwsid,12),
                                                                         "Name" = rep(system_name,12),
                                                                         "Ownership" = rep(ownership,12))

#5 add group 1 so that data will be considered together
ggplot(withdrawls, aes(x = Date, y = max_mgd, group = 1))+
  geom_line()+
  geom_smooth(method="loess", se=FALSE)+
  labs(title = "Maximum Water Usage by Month", x = "Date", y = "Water Usage (million gallons per day)")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
#scraping function
scrape_page <- function(the_year, the_pwsid){
  the_page <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', the_pwsid, '&year=', the_year))

#set element address
pwsid_tag <- "td tr:nth-child(1) td:nth-child(5)"
max_daily_tag <- "th~ td+ td"
month_tag <- ".fancy-table:nth-child(31) tr+ tr th"
system_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"

#scrape the data
the_pwsid <- the_page %>% html_nodes(pwsid_tag) %>% html_text()
the_max_daily <- the_page %>% html_nodes(max_daily_tag) %>% html_text()
the_month <- the_page %>% html_nodes(month_tag) %>% html_text()
system_name <- the_page %>% html_nodes(system_tag) %>% html_text()

#convert to a dataframe
the_withdrawals <- data.frame("Month" = c('Jan','May', 'Sep', 'Feb', 'Jun', 'Oct', 'Mar', 'Jul', 	
'Nov', 'Apr', 'Aug', 'Dec'),
                              "Year" = rep(the_year,12),
                              "MGD" = as.numeric(the_max_daily)) %>%
  mutate(system_name = !!system_name,
    Date = my(paste(Month,"-",Year)))
return(the_withdrawals)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
#run function
Durham_2015 <- scrape_page(2015,'03-32-010')
view(Durham_2015)

#plot
ggplot(Durham_2015) + 
  geom_line(aes(x = Date, y = MGD))+
  labs(title = "Water Usage Data by Month in 2015", x = "Date", y = "Withdrawal (mgd)")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
Ashville_2015 <- scrape_page(2015,'01-11-010')
view(Ashville_2015)

Ashville_Durham_2015 <- rbind(Ashville_2015, Durham_2015)

ggplot(Ashville_Durham_2015, aes(x = Date, y = MGD, color = system_name)) +
  geom_line() + 
  geom_smooth(method="loess", se=FALSE) +
  labs(title = "Water Usage Data by Month in 2015 in Ashville v. Durham",
       y="Withdrawal (mgd)",
       x="Date")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
A_PWSID <- '01-11-010'
  
Asheville_2010_2021 <- map2(2010:2021, A_PWSID, scrape_page) %>%
  bind_rows()

ggplot(Asheville_2010_2021, aes(x = Date, y = MGD))+
  geom_point()+
  geom_smooth(method="loess", se=FALSE)+
  labs(title = "Water Usage Data by Month in Ashville from 2010 to 2021",
       y="Withdrawal (mgd)",
       x="Date")

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: yes, more water is being used overtime since 2015.
>
